---
title: "Analysis of the New York Housing Market"
format: html
---


```{r}

library(dplyr)
library(ggplot2)
library(plotly)
library(tidyr)
library(hexbin)
library(RColorBrewer)
library(forcats)
library(crosstalk)
library(ggiraph)

library(survey)

library(srvyr)

library(tidycensus)

library(sf)

library(jsonlite)

library(tigris)

library(stringr)

library(leaflet)
library(htmlwidgets)
library(htmltools)
library(gt)
library(DT)


```




# New York Housing data Review


```{r}


saveRDS(ny_housing, "ny_housing.rds")


```


```{r}


ny_housing<-readRDS("ny_housing.rds")

```



```{r}



ny_housing |> glimpse()


```



```{r}



ny_housing |> summary()


```



```{r}

ny_housing |> 
    select(STATE, MAIN_ADDRESS, FORMATTED_ADDRESS, LOCALITY) |> 
    head(20)


```




The address fields are not consistant on the city or town names.  In some instances the city name is actually a neighborhood within a city or New York City borough. 

We will use the stringr package and regular expressions to extract the zip code which then can be used to determine the city. 





```{r}


ny_housing_2<-ny_housing |> 
    mutate(zip_cde=str_extract(MAIN_ADDRESS,"\\d{5}"))




```


```{r}


ny_housing_2 |> 
    select(zip_cde) |> 
    head(20)



```




```{r}


glimpse(ny_housing_2)


```






```{r}

ny_housing_2 |> 
    select(zip_cde) |> 
    head(20)


```


```{r}


saveRDS(ny_housing_2,"ny_housing_2.rds")


```


```{r}



ny_housing_2 <- readRDS("ny_housing_2.rds")


```









```{r}






```





# Cenus







```{r}



Sys.getenv("CENSUS_API_KEY")

```



```{r}



census_2020<-load_variables(2020,"pl",cache=TRUE)


```

## Longitude/Latitude 


For our analysis we will require census tract codes that match our NY housing data. We will use the longitude and latitude columns in the dataset for this.  The first step is to create a data frame which includes only the latitude and longitude columns. 

```{r}



long_lat <- ny_housing_2 |> 
    select(LATITUDE, LONGITUDE)


```


```{r}


long_lat |> glimpse()



```



Now that we have our long_lat data frame we will use the call_geolocator_latlon function from the Tigris library.  The function submits a request to the census bureau api with a new column of census_code specified. 

```{r}



long_lat$census_code <- apply(long_lat, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))




```


```{r}


long_lat_data <-long_lat



```


```{r}

saveRDS(long_lat_data,"long_lat_data.rds")



```



```{r}


long_lat_data <-readRDS("long_lat_data.rds")

```


We can observe that the data frame returned from the cenus api now includes the cenus_code column. 


```{r}



long_lat_data |> head()


```

```{r}


long_lat_data |> glimpse()


```


The 15 digit code returnd  from the census api is made up of serveral codes put together.  The first two digits are the state code.  The next three digits are the county code. The next six digits are the census tract code.  The last four digits are the census block code. 


```{r}



ny_tracts <- tracts("NY", year = 2024)


```

```{r}


class(ny_tracts)



```

```{r}


ny_tracts |>  head()


```


```{r}


census_codes<-long_lat_data



```



We will use the substr function to break up the census code into its individual components.  THe census tract will first be extracted. 


```{r}



census_codes$census_tract<-substr(census_codes$census_code,6,11)


```



Three digit county code

```{r}



census_codes$county <-substr(census_codes$census_code,3,5)



```



four digit census block code
 


```{r}

census_codes$block <-substr(census_codes$census_code,12,15)


```

#GEOID column



```{r}



census_codes$GEOID <-substr(census_codes$census_code,1,11)


```



```{r}


saveRDS(census_codes, "census_codes.rds")



```


```{r}


census_codes <- readRDS("census_codes.rds")


```


```{r}



census_codes |> head()


```

```{r}


census_codes |>  glimpse()



```



The output displays our added census tract, count, block codes, and GEOID. 






```{r}


census_geo <- census_codes |> 
    select(-LATITUDE,-LONGITUDE)


```


```{r}



saveRDS(census_geo,"census_geo.rds")


```


```{r}


census_geo <- readRDS("census_geo.rds")

```

```{r}

census_geo |> head()



```

```{r}


census_geo |>  glimpse()

```




```{r}


ny_housing_census <- cbind(ny_housing_2, census_geo)


```



```{r}



ny_housing_census <- ny_housing_census |> 
    select(GEOID,census_code,TYPE, PRICE, BEDS, BATH, PROPERTYSQFT, LATITUDE, LONGITUDE, zip_cde, census_tract, county, block)


```



```{r}



saveRDS(ny_housing_census, "ny_housing_census.rds")


```


```{r}


ny_housing_census <- readRDS("ny_housing_census.rds")


```

```{r}

ny_housing_census |> head()


```




```{r}


ny_housing_census |> glimpse()



```


```{r}



ny_housing_census |> summary()

```

## Census tract Demographic Data




```{r}



load_variables(2024, "acs5")


```


Requesting census data from different tables can cause errors. To prevent this, we will request data separatley for the two tables. 


```{r}


variables_to_get_1 <- c(
    median_value = "B25077_001",
    total_population = "B01003_001",
    median_age = "B01002_001",
    median_year_built = "B25037_001"
   

) 






```


```{r}




variables_to_get_2 <- c(
    median_income = "DP03_0062",
    pct_college = "DP02_0068P",
    pct_foreign_born = "DP02_0094P",
    pct_white = "DP05_0077P",
    percent_ooh = "DP04_0046P"

) 


```


```{r}


ny_tract_info_1 <- get_acs(
    geography = "tract",
    variables = variables_to_get_1,
    state = "NY",
    #geometry = TRUE,
    output="wide",
    year = 2024
) #|> 
    #select(-NAME)



```


```{r}



ny_tract_info_1 |> head()



```

```{r}



ny_tract_info_1 |> glimpse()


```





```{r}


saveRDS(ny_tract_info_1,"ny_tract_info_1.rds")



```

```{r}



ny_tract_info_1 <- readRDS("ny_tract_info_1.rds")


```

```{r}


ny_tract_info_2 <- get_acs(
    geography = "tract",
    variables = variables_to_get_2,
    state = "NY",
    #geometry = TRUE,
    output="wide",
    year = 2024
) #|> 
    #select(-NAME)




```



```{r}


ny_tract_info_2 <- ny_tract_info_2 |> 
    select(-geometry)



```


```{r}



saveRDS(ny_tract_info_2,"ny_tract_info_2.rds")



```






```{r}



ny_tract_info_2 <- readRDS("ny_tract_info_2.rds")


```


```{r}

ny_tract_info_2 |> glimpse()



```


```{r}

ny_tract_geometry <- ny_tract_info_1 |> 
    select(GEOID, geometry)


```


```{r}


ny_tract_geometry |> glimpse()


```


```{r}


saveRDS(ny_tract_geometry, "ny_tract_geometry.rds")


```


```{r}



ny_tract_geometry <- readRDS("ny_tract_geometry.rds")


```


```{r}


ny_tract_data <- ny_tract_info_1 |> 
    full_join(ny_tract_info_2, by="GEOID")


```

```{r}

st_join()


```





```{r}


ny_tract_data |> glimpse()


```


```{r}


ny_tract_data <-ny_tract_data |> 
    select(-NAME.y)


```

```{r}

ny_tract_data <- ny_tract_data |> 
    rename(NAME=NAME.x)

```



```{r}

saveRDS(ny_tract_data, "ny_tract_data.rds")


```


```{r}



ny_tract_data <-readRDS("ny_tract_data.rds")


```

```{r}

ny_tract_data |> summary()



```


```{r}



ny_tract_data |> glimpse()


```



```{r}


ny_tract_geom <- st_join(ny_tract_data,ny_tract_geometry)


```

## Feature Engineering

```{r}


library(units)


```


```{r}


ny_tract_data_2 <- ny_tract_data |> 
    mutate(pop_density = as.numeric(set_units(total_populationE / st_area(.),"1/km2")),
           median_structure_age = 2022 - median_year_builtE) |> 
    select(!ends_with("M")) |> 
    rename_with(.fn = ~str_remove(.x, "E$"))



```



```{r}


load_variables(2020,"acs5")



```