---
title: "Analysis of the New York Housing Market"
format: html
---


```{r}

library(dplyr)
library(ggplot2)
library(plotly)
library(tidyr)
library(hexbin)
library(RColorBrewer)
library(forcats)
library(crosstalk)
library(ggiraph)

library(survey)

library(srvyr)

library(tidycensus)

library(sf)

library(jsonlite)

library(tigris)

library(stringr)

library(leaflet)
library(htmlwidgets)
library(htmltools)
library(gt)
library(DT)
library(GGally)
library(lme4)
library(lmerTest)
library(rstanarm)


```




# New York Housing data Review


```{r}


saveRDS(ny_housing, "ny_housing.rds")


```


```{r}


ny_housing<-readRDS("ny_housing.rds")

```



```{r}



ny_housing |> glimpse()


```



```{r}



ny_housing |> summary()


```



```{r}

ny_housing |> 
    select(STATE, MAIN_ADDRESS, FORMATTED_ADDRESS, LOCALITY) |> 
    head(20)


```




The address fields are not consistant on the city or town names.  In some instances the city name is actually a neighborhood within a city or New York City borough. 

We will use the stringr package and regular expressions to extract the zip code which then can be used to determine the city. 





```{r}


ny_housing_2<-ny_housing |> 
    mutate(zip_cde=str_extract(MAIN_ADDRESS,"\\d{5}"))




```



```{r}


saveRDS(ny_housing_2,"ny_housing_2.rds")


```



```{r}


ny_housing_2 <- readRDS("ny_housing_2.rds")




```


```{r}


ny_housing_2 |> 
    select(zip_cde) |> 
    head(20)



```




```{r}


glimpse(ny_housing_2)


```






```{r}

ny_housing_2 |> 
    head(20)


```





```{r}






```

```{r}







```









```{r}






```












```{r}



Sys.getenv("CENSUS_API_KEY")

```



```{r}



census_2020<-load_variables(2020,"pl",cache=TRUE)


```

## Longitude/Latitude 


For our analysis we will require census tract codes that match our NY housing data. We will use the longitude and latitude columns in the dataset for this.  The first step is to create a data frame which includes only the latitude and longitude columns. 

```{r}



long_lat <- ny_housing_2 |> 
    select(LATITUDE, LONGITUDE)


```


```{r}


long_lat |> glimpse()



```



Now that we have our long_lat data frame we will use the call_geolocator_latlon function from the Tigris library.  The function submits a request to the census bureau api with a new column of census_code specified. 

```{r}



long_lat$census_code <- apply(long_lat, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))




```


```{r}


long_lat_data <-long_lat



```


```{r}

saveRDS(long_lat_data,"long_lat_data.rds")



```



```{r}


long_lat_data <-readRDS("long_lat_data.rds")

```


We can observe that the data frame returned from the cenus api now includes the cenus_code column. 


```{r}



long_lat_data |> head()


```

```{r}


long_lat_data |> glimpse()


```


The 15 digit code returnd  from the census api is made up of serveral codes put together.  The first two digits are the state code.  The next three digits are the county code. The next six digits are the census tract code.  The last four digits are the census block code. 


```{r}



ny_tracts <- tracts("NY", year = 2024)


```

```{r}


class(ny_tracts)



```

```{r}


ny_tracts |>  head()


```


```{r}


census_codes<-long_lat_data



```



We will use the substr function to break up the census code into its individual components.  THe census tract will first be extracted. 


```{r}



census_codes$census_tract<-substr(census_codes$census_code,6,11)


```



Three digit county code

```{r}



census_codes$county <-substr(census_codes$census_code,3,5)



```



four digit census block code
 


```{r}

census_codes$block <-substr(census_codes$census_code,12,15)


```

GEOID column



```{r}



census_codes$GEOID <-substr(census_codes$census_code,1,11)


```



```{r}


saveRDS(census_codes, "census_codes.rds")



```


```{r}


census_codes <- readRDS("census_codes.rds")


```


```{r}



census_codes |> head()


```

```{r}


census_codes |>  glimpse()



```



The output displays our added census tract, count, block codes, and GEOID. 






```{r}


census_geo <- census_codes |> 
    select(-LATITUDE,-LONGITUDE)


```


```{r}



saveRDS(census_geo,"census_geo.rds")


```


```{r}


census_geo <- readRDS("census_geo.rds")

```

```{r}

census_geo |> head()



```

```{r}


census_geo |>  glimpse()

```




```{r}


ny_housing_census <- cbind(ny_housing_2, census_codes)


```



```{r}



ny_housing_census <- ny_housing_census |> 
    select(GEOID,census_code,TYPE, PRICE, BEDS, BATH, PROPERTYSQFT, LATITUDE, LONGITUDE, zip_cde, census_tract, county, block)


```



```{r}



saveRDS(ny_housing_census, "ny_housing_census.rds")


```


```{r}


ny_housing_census <- readRDS("ny_housing_census.rds")


```

```{r}

ny_housing_census |> head()


```




```{r}


ny_housing_census |> glimpse()



```


```{r}



ny_housing_census |> summary()

```



```{r}



ny_housing_census |> 
    count(TYPE) |> 
    arrange(desc(n))



```


For TYPE, certain categories (Pending, Contingent, For Sale, Foreclsure, Coming Soon) offer little information on what the property is. Additionally, "Mobile  house for sale" has a count of only one. These categories will be removed.  


```{r}


ny_housing_census_2 <- ny_housing_census |> 
    filter(TYPE %in% c("Co-op for sale", "House for sale", "Condo for sale", "Multi-family home for sale", "Townhouse for sale", "Land for sale", "Condop for sale"))


```



We will create artifical id's for each property.  This will be useful during our modeling. 


```{r}



ny_housing_census_2 <- ny_housing_census_2 |> 
    mutate(property_id = row_number())


```


The TYPE category "Condop for sale" will be merged into "Condo for sale" 


```{r}



ny_housing_census_2 <- ny_housing_census_2 |> 
    mutate(TYPE = recode(TYPE, "Condop for sale" = "Condo for sale"))




```


```{r}


ny_housing_census_2 <- ny_housing_census_2 |> 
    mutate(BATH = round(BATH,0)) |> 
    mutate(PROPERTYSQFT = round(PROPERTYSQFT,0))


```



```{r}


ny_housing_census_2 <- ny_housing_census_2 |> 
    select(-census_tract)



```



```{r}


ny_housing_census_2 <- ny_housing_census_2 |> 
    mutate(property_id = as.character(property_id))

```


```{r}



saveRDS(ny_housing_census_2, "ny_housing_census_2.rds")


```


```{r}


ny_housing_census_2 <- readRDS("ny_housing_census_2.rds")


```


```{r}

ny_housing_census_2 |> 
    glimpse()


```


```{r}


ny_housing_census_2 |> 
    count(TYPE) |> 
    arrange(desc(n))

```

The feature TYPE has been reduced to six categories. 



```{r}


ny_housing_census_2 |> 
    count(TYPE) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))



```

Above is the percentage breakdown of propery type.





```{r}



ny_housing_census_2 |> 
    select(property_id) |>
     head()


```






```{r}


dim(ny_housing_census_2)



```

Our dataset now has 4,433 observations. 


```{r}



```





```{r}


ny_housing_census_2 |> 
    count(county) |> 
    arrange(desc(n))


```


There are five counties in our data that have properties for sale. 




```{r}



ny_housing_census_2 |> 
    count(TYPE, GEOID) |>
    slice_max(n,n=10) |> 
    arrange(desc(n))



```

The top 10 tracts by count of property for sale. 


```{r}


options(scipen = 999)


```


```{r}

ny_housing_census_2 |> 
    select(PRICE, BEDS, BATH, PROPERTYSQFT) |> 
    summary()


```




```{r}


ny_housing_census_2 |> glimpse()


```


```{r}


county_level<-ny_housing_census_2 |> 
    select(GEOID, TYPE, PRICE, BEDS, BATH, PROPERTYSQFT, county, property_id)


```


```{r}


county_level$GEOID <- substr(county_level$GEOID,1,5)


```


```{r}

saveRDS(county_level, "county_level.rds")


```

```{r}

county_level <- readRDS("county_level.rds")


```


```{r}


property_county<-county_level |> 
    count(GEOID, name = "county_property_cnt")


```












```{r}


write.csv(ny_housing_census, "ny_housing_census.csv")



```





# Census 




```{r}



load_variables(2024, "acs5")


```


## Tract Demographic Data


Requesting census data from different tables can cause errors. To prevent this, we will request data separatley for the two tables. 


```{r}


variables_to_get_1 <- c(
    median_value = "B25077_001",
    total_population = "B01003_001",
    median_age = "B01002_001",
    median_year_built = "B25037_001"
   

) 






```


```{r}




variables_to_get_2 <- c(
    median_income = "DP03_0062",
    pct_college = "DP02_0068P",
    pct_foreign_born = "DP02_0094P",
    pct_white = "DP05_0077P",
    percent_ooh = "DP04_0046P"

) 


```


```{r}


ny_tract_info_1 <- get_acs(
    geography = "tract",
    variables = variables_to_get_1,
    state = "NY",
    #geometry = TRUE,
    output="wide",
    year = 2024
) #|> 
    #select(-NAME)



```


```{r}



ny_tract_info_1 |> head()



```

```{r}



ny_tract_info_1 |> glimpse()


```





```{r}


saveRDS(ny_tract_info_1,"ny_tract_info_1.rds")



```

```{r}



ny_tract_info_1 <- readRDS("ny_tract_info_1.rds")


```


```{r}


ny_tract_info_1 |> 
    glimpse()

```

```{r}


ny_tract_info_2 <- get_acs(
    geography = "tract",
    variables = variables_to_get_2,
    state = "NY",
    #geometry = TRUE,
    output="wide",
    year = 2024
) #|> 
    #select(-NAME)




```



```{r}


ny_tract_info_2 <- ny_tract_info_2 |> 
    select(-geometry)



```


```{r}



saveRDS(ny_tract_info_2,"ny_tract_info_2.rds")



```






```{r}



ny_tract_info_2 <- readRDS("ny_tract_info_2.rds")


```


```{r}

ny_tract_info_2 |> glimpse()



```


```{r}

ny_tract_geometry <- ny_tract_info_1 |> 
    select(GEOID, geometry)


```


```{r}


ny_tract_geometry |> glimpse()


```


```{r}


saveRDS(ny_tract_geometry, "ny_tract_geometry.rds")


```


```{r}



ny_tract_geometry <- readRDS("ny_tract_geometry.rds")


```


```{r}


ny_tract_geometry |> head()


```


```{r}



class(ny_tract_geometry)


```

```{r}

ny_tract_cls <- ny_tract_geometry |> 
    st_transform(5070) 


```


```{r}


saveRDS(ny_tract_cls, "ny_tract_cls.rds")



```


```{r}


ny_tract_cls <- readRDS("ny_tract_cls.rds")


```


```{r}


class(ny_tract_cls)


```

```{r}



ny_tract_cls |> head()



```


```{r}


ny_tract_data <- ny_tract_info_1 |> 
    full_join(ny_tract_info_2, by="GEOID")


```




```{r}







```


```{r}


ny_tract_data |> glimpse()


```


```{r}


ny_tract_data <-ny_tract_data |> 
    select(-NAME.y)


```

```{r}

ny_tract_data <- ny_tract_data |> 
    rename(NAME=NAME.x)

```



```{r}


saveRDS(ny_tract_data, "ny_tract_data.rds")




```



```{r}



ny_tract_data <-readRDS("ny_tract_data.rds")



```



```{r}



ny_tract_data |> glimpse()

```


```{r}



ny_tract_data_2 <- ny_tract_data |> 
    separate(NAME, into=c("census_tract", "county", "state"), sep="; ")


```


```{r}



ny_tract_data_2 |> glimpse()



```




```{r}



ny_tract_data_2 <- ny_tract_data_2 |> 
    select(-state)



```

```{r}





```



```{r}

saveRDS(ny_tract_data_2, "ny_tract_data_2.rds")


```


```{r}



ny_tract_data_2 <-readRDS("ny_tract_data_2.rds")


```

```{r}

ny_tract_data_2 |> summary()



```


```{r}



ny_tract_data_2 |> glimpse()


```



```{r}



ny_tract_geom<-ny_tract_cls |> 
    full_join(ny_tract_data_2, by = "GEOID")


```




```{r}



ny_tract_geom<-ny_tract_geometry |> 
    full_join(ny_tract_data_2, by = "GEOID")



```




```{r}

saveRDS(ny_tract_geom,"ny_tract_geom.rds")


```


```{r}


ny_tract_geom <- readRDS("ny_tract_geom.rds")


```

```{r}

ny_tract_geom |> glimpse()


```

```{r}


ny_tract_geom |> summary()


```



```{r}


#ny_tract_geom <- st_join(ny_tract_data,ny_tract_geometry)


```

## Feature Engineering

```{r}


library(units)


```


```{r}


ny_tract_geom_2 <- ny_tract_geom |> 
    mutate(pop_density = as.numeric(set_units(total_populationE / st_area(geometry),"1/km2")),
           median_structure_age = 2022 - median_year_builtE) |> 
    select(!ends_with("M")) |> 
    rename_with(.fn = ~str_remove(.x, "E$"))



```



```{r}



ny_tract_geom_2 <- ny_tract_geom_2 |> 
    mutate(pop_density = round(pop_density,2))



```


```{r}


ny_tract_geom_2 <- ny_tract_geom_2 |> 
    rename(county_name = county)



```



```{r}


saveRDS(ny_tract_geom_2 , "ny_tract_geom_2.rds")


```



```{r}



ny_tract_geom_2  <- readRDS("ny_tract_geom_2.rds")


```



```{r}


ny_tract_geom_2 |> head(20)


```


```{r}



ny_tract_geom_2 |> 
    summary()


```



```{r}


ny_tract_geom_2 |> 
    filter(census_tract == "Census Tract 102") |> 
    head()


```




```{r}

ny_tract_geom_2 |> 
    select(census_tract) |> 
    count(census_tract) |> 
    arrange(desc(n))


```






```{r}



ny_tract_geom_2 |> 
    glimpse()




```

## Join Housing and Tract Data



```{r}





ny_census_combined <- ny_tract_geom_2 |> 
    inner_join(ny_housing_census_2, join_by(GEOID)) |> 
    select(-census_code, -zip_cde, -block)



```


```{r}


saveRDS(ny_census_combined, "ny_census_combined.rds")

```


```{r}

ny_census_combined <- readRDS("ny_census_combined.rds")


```




# Data Cleaning

## Missing Values


```{r}



ny_census_combined |> glimpse()

```

```{r}

ny_census_combined |>  summary()


```



There are a number of features with missing values.  Lets inspect these missing values. 



### Median year built

```{r}



ny_census_combined |> 
    filter(is.na(median_year_built)) |> 
    summary()



```


```{r}




ny_census_combined |> 
    filter(is.na(median_year_built)) |> 
    head()



```

There are three Census tracts which are missing demographic data.  We will remove these missing values from the dataset. 



```{r}



ny_census_combined_2 <- ny_census_combined |> 
    filter(!is.na(median_year_built))



```


```{r}

saveRDS(ny_census_combined_2, "ny_census_combined_2.rds")


```

```{r}


ny_census_combined_2 <- readRDS("ny_census_combined_2.rds")


```




```{r}


ny_census_combined_2 |> 
    filter(is.na(median_year_built)) |> 
    summary()


```

The three census tracts with missing demographic data have been removed. 


### Medium Income


```{r}




ny_census_combined_2 |> 
    summary()




```


There are missing values for median_income.  

```{r}


ny_census_combined_2 |> 
    filter(is.na(median_income)) |> 
    head(17)


```


```{r}

ny_census_combined_2 |> 
    filter(is.na(median_income)) |> 
    tail(10)


```

The tracts with missing median_income values have pct_white populations that are low or zero.  AS such it would be best for our analyis to retain these rows.  We will replace missing values with the median of the tract median_income.   


```{r}






```

Census tracts 102 and 96 within New York County

We'll use median income of all tracts within the county the tract witgh the missing values reside in. 

```{r}


ny_tract_geom_2 |> 
    filter(county_name == "New York County") |> 
    filter(!is.na(median_income)) |> 
    summarise(median_tract_income = median(median_income))



```


```{r}



ny_census_combined_3 <- ny_census_combined_2 |> 
    mutate(median_income = ifelse(county_name == "New York County" & is.na(median_income), 121760.5,median_income ))


```



```{r}


ny_census_combined_3 |> 
    filter(is.na(median_income)) |> 
    head(17)


```

As New York County includes tracts 102 and 96, we can observe those those tracts no longer have missing values for median income. 






Census Tract 226.02 Richmond County

```{r}

ny_tract_geom_2 |> 
    filter(county_name == "Richmond County") |> 
    filter(!is.na(median_income)) |> 
    summarise(median_tract_income = median(median_income))



```





census tract 226.02



```{r}

ny_census_combined_3 |> 
    filter(census_tract == "Census Tract 102") |> 
    filter(is.na(median_income)) 


```


```{r}


ny_census_combined_3 <- ny_census_combined_3 |> 
    mutate(median_income = ifelse(county_name == "Richmond County" & is.na(median_income), 99305,median_income ))

```



Census Tract 932  Kings County 

```{r}



ny_tract_geom_2 |> 
    filter(county_name == "Kings County") |> 
    filter(!is.na(median_income)) |> 
    summarise(median_tract_income = median(median_income))



```



```{r}



ny_census_combined_3 <- ny_census_combined_3 |> 
    mutate(median_income = ifelse(county_name == "Kings County" & is.na(median_income), 79773,median_income ))


```


Census Tract 231  Bronx County



```{r}


ny_tract_geom_2 |> 
    filter(county_name == "Bronx County") |> 
    filter(!is.na(median_income)) |> 
    summarise(median_tract_income = median(median_income))


```


```{r}


ny_census_combined_3 <- ny_census_combined_3 |> 
    mutate(median_income = ifelse(county_name == "Bronx County" & is.na(median_income), 50243,median_income ))


```


Census Tract 536.01 Queens County

```{r}


ny_tract_geom_2 |> 
    filter(county_name == "Queens County") |> 
    filter(!is.na(median_income)) |> 
    summarise(median_tract_income = median(median_income))

```


```{r}


ny_census_combined_3 <- ny_census_combined_3 |> 
    mutate(median_income = ifelse(county_name == "Queens County" & is.na(median_income), 88882,median_income ))


```


```{r}


saveRDS(ny_census_combined_3, "ny_census_combined_3.rds")




```



```{r}

ny_census_combined_3 <- readRDS("ny_census_combined_3.rds")


```


```{r}



ny_census_combined_3 |> 
    filter(is.na(median_income)) |> 
    head()



```

```{r}


ny_census_combined_3 |> 
    summary()


```






All missing values for median income have been filled in (impouted).  The last feature to inspect is median_value. 

### Median Value


```{r}




ny_census_combined_3 |> 
    filter(is.na(median_value)) |> 
    head(10)


```


```{r}



ny_census_combined_3 |> 
    filter(is.na(median_value)) |> 
    tail(10)


```





 Census Tracts 391, 215.02 , 266.01, 375.04, 173, 419 Bronx County








```{r}


ny_tract_geom_2 |> 
    filter(county_name == "Bronx County") |> 
    filter(!is.na(median_value)) |> 
    summarise(median_tract_value = median(median_value))



```




```{r}



ny_census_combined_4 <- ny_census_combined_3 |> 
    mutate(median_value = ifelse(county_name == "Bronx County" & is.na(median_value), 607050,median_value ))



```




Census Tracts 505, 596, 575 Queens County





```{r}


ny_tract_geom_2 |> 
    filter(county_name == "Kings County") |> 
    filter(!is.na(median_value)) |> 
    summarise(median_tract_value = median(median_value))


```



```{r}


ny_census_combined_4 <- ny_census_combined_4 |> 
    mutate(median_value = ifelse(county_name == "Kings County" & is.na(median_value), 956900,median_value ))


```



Census Tracts 96, 172, 99.0293 New York County


```{r}


ny_tract_geom_2 |> 
    filter(county_name == "New York County") |> 
    filter(!is.na(median_value)) |> 
    summarise(median_tract_value = median(median_value))



```

```{r}


ny_census_combined_4 <- ny_census_combined_4 |> 
    mutate(median_value = ifelse(county_name == "New York County" & is.na(median_value), 1078250,median_value ))

```

Tracts 85, 938, 83, 1227.02, 1133, 779.08, 155, 240 Queens County


```{r}


ny_tract_geom_2 |> 
    filter(county_name == "Queens County") |> 
    filter(!is.na(median_value)) |> 
    summarise(median_tract_value = median(median_value))

```



```{r}


ny_census_combined_4 <- ny_census_combined_4 |> 
    mutate(median_value = ifelse(county_name == "Queens County" & is.na(median_value), 746350,median_value ))



```



```{r}


ny_census_combined_4 <- ny_census_combined_4 |> 
    mutate(property_id = as.character(property_id))

```


```{r}


saveRDS(ny_census_combined_4 , "ny_census_combined_4.rds")


```


```{r}


ny_census_combined_4  <- readRDS("ny_census_combined_4.rds")


```

All missing values have been removed or imputed. 


```{r}


ny_census_combined_4 |> 
    summary()


```


```{r}

ny_census_combined_4 |> glimpse()


```

# Visualization



## Features


```{r}



sqr_ft_box<-ny_housing_census_2 %>% 
  ggplot(aes(x=TYPE, y=PROPERTYSQFT, fill=TYPE))+
  geom_boxplot()+
  ggtitle("Property Square Feet by Property Type")+
  theme(legend.position = "none",plot.title = element_text(color="blue", size=14, face="bold.italic", hjust=0.5))+
  theme(axis.ticks.x=element_blank(),legend.position = "none", axis.text.x=element_text(angle=70, hjust=1),
        axis.title.x=element_blank(),
    axis.title.y=element_blank())



```


```{r}


sqr_ft_box

```



```{r}




bath_box<-ny_housing_census_2 %>% 
  ggplot(aes(x=TYPE, y=BATH, fill=TYPE))+
  geom_boxplot()+
  ggtitle("Number of Bath Rooms  by Property Type")+
  theme(legend.position = "none",plot.title = element_text(color="blue", size=14, face="bold.italic", hjust=0.5))+
  theme(axis.ticks.x=element_blank(),legend.position = "none", axis.text.x=element_text(angle=70, hjust=1),
        axis.title.x=element_blank(),
    axis.title.y=element_blank())



```

```{r}


bath_box

```



```{r}


bed_box<-ny_housing_census_2 %>% 
  ggplot(aes(x=TYPE, y=BEDS, fill=TYPE))+
  geom_boxplot()+
  ggtitle("Number of Bed Rooms  by Property Type")+
  theme(legend.position = "none",plot.title = element_text(color="blue", size=14, face="bold.italic", hjust=0.5))+
  theme(axis.ticks.x=element_blank(),legend.position = "none", axis.text.x=element_text(angle=70, hjust=1),
        axis.title.x=element_blank(),
    axis.title.y=element_blank())



```


```{r}

bed_box


```


```{r}



price_hist <- ggplot(ny_housing_census_2, aes(x=PRICE)) +
    geom_histogram(bins=30) +
    scale_x_continuous() 


```

```{r}

price_hist


```


Our historgram for the PRICE feature shows all our observations falling at zero.  This does not seem correct.  LEt's review a summary of the PRICE feature. 


```{r}


ny_housing_census_2 |> 
    select(PRICE) |> 
    summary()


```

There is a dramatic range of values from a minimum of 5800 to a maximum of 2,147,483,647




```{r}


ny_housing_census_2 |> 
    filter(PRICE > 100000000)



```

We have two properties with a price over 100 million.  This might be affecting a plot.  Let's first remove the two observations over 100 million and plot the histogram. 


```{r}

ny_housing_census_2 |> 
    filter(PRICE < 100000000) |> 
    ggplot(aes(x=PRICE)) +
    geom_histogram(bins=60) +
    scale_x_continuous()




```



That helped though there is right scew to our plot.  Instead of removing values, we'll apply a logarithmic transformation to the PRICE feature. 


```{r}

log_price_hist <- ggplot(ny_housing_census_2, aes(x=log(PRICE))) +
    geom_histogram(bins=30) +
    scale_x_continuous() 


```




```{r}

log_price_hist


```



Applying the logarithmic tranformation to PRICE gives us a plot closer to the normal distribution (bell shaped).  






```{r}


sqft_hist <- ggplot(ny_housing_census_2, aes(x=PROPERTYSQFT)) +
    geom_histogram(bins=100) +
    scale_x_continuous() 


```

```{r}

sqft_hist

```

The property squarefeet histogram has a scew to teh right.  bWE'll try apllying a logarithmic transformation to this feature. 


```{r}



log_sqft_hist <- ggplot(ny_housing_census_2, aes(x=log(PROPERTYSQFT))) +
    geom_histogram(bins=100) +
    scale_x_continuous() 


```


```{r}
log_sqft_hist


```


The log transformation gave us closer to a normal distribution though we see that there are many values around the mean. 


```{r}


ny_housing_census_2 |> 
    select(PROPERTYSQFT) |> 
    summary()


```

 ```{r}
 

ny_housing_census_2 |> 
    select(PROPERTYSQFT) |> 
    filter(PROPERTYSQFT >= 2184 & PROPERTYSQFT <=2198) |> 
    


 ```


 Roughly Thirty Five Percent of our obesrvations fall between PROPERTY Square Feet of 2184 and 2198. 


```{r}




corr_plot <- ny_housing_census_2 |> 
    select(PRICE, PROPERTYSQFT, BEDS, BATH) |> 
    ggscatmat(corMethod = "pearson", alpha = 0.2)

```



```{r}


corr_plot


```





## Maps


```{r}


ny_county_shape <- counties(state="NY", cb=FALSE, year=2024)


```


```{r}


saveRDS(ny_county_shape, "ny_county_shape.rds")


```


```{r}

ny_county_shape <- readRDS("ny_county_shape.rds")


```


```{r}


county_level_2 <- ny_county_shape |> 
    inner_join(county_level, by = "GEOID")


```

```{r}

county_level_2 <- county_level_2 |> 
    select(GEOID, TYPE, PRICE, BEDS, BATH, property_id, geometry)



```


```{r}


saveRDS(county_level_2, "county_level_2.rds")



```

```{r}


county_level_2 <- readRDS("county_level_2.rds")

```



```{r}


ggplot()+
    geom_sf(data=ny_county_shape, fill=NA)+
    geom_sf(data=county_level_2, color="red", size=1)+
    theme_bw()



```


From this plot it appears that all properties for sale are in counties that include all five boroughs of New York City.  Let's create a map that focuses on this area. 

```{r}



ny_county_shape_2 <- ny_county_shape |> 
    filter(NAME %in% c("Bronx", "Kings", "New York", "Queens", "Richmond" ))


```


```{r}


saveRDS(ny_county_shape_2, "ny_county_shape_2.rds")


```

```{r}

ny_county_shape_2 <- readRDS("ny_county_shape_2.rds")


```

```{r}

st_crs(ny_county_shape_2)

```


```{r}



ny_county_epsg4326 <- ny_county_shape_2 |> 
    st_transform(crs=4326)

```



```{r}


saveRDS(ny_county_epsg4326, "ny_county_epsg4326.rds")

```


```{r}



ggplot()+
    geom_sf(data=ny_county_epsg4326, fill=NA)+
    geom_sf(data=county_level_2, color="red", size=1)+
    theme_bw()


```



```{r}

ny_tracts <- tracts(state="NY", cb=FALSE, year=2024)


```



```{r}

saveRDS(ny_tracts, "ny_tracts.rds")


```



```{r}


ny_tracts <- readRDS("ny_tracts.rds")


```






```{r}



ny_tracts <- readRDS("ny_tracts.rds")


```


This plot displays how the geographical size of census tracts can vary. 


```{r}


ny_tracts_plot <- ggplot(ny_tracts)+
    geom_sf()+
    theme_void()



```


```{r}






```


```{r}



ny_tracts <- tracts(state="NY", cb=FALSE, year=2024)


```


We're going to create a map displaying the median home value of the census tract. First, we are going to create a feature for the property count in each census tract. This will be used for hovering data on our map.   

```{r}


property_tract_count<-ny_housing_census_2 |> 
    count(GEOID, name = "tract_property_count")
    


```



```{r}

saveRDS(property_tract_count, "property_tract_count.rds")


```


```{r}

property_tract_count <- readRDS("property_tract_count.rds")


```



```{r}


property_tract_count |> 
    glimpse()


```

```{r}


property_tract_count |> 
    slice_max(tract_property_count, n=20) 
    
   


```


```{r}

property_tract_count |> 
    summary()



```


```{r}


class(property_tract_count)



```


We will now add our tract property count feature to our combined data. 


```{r}


ny_census_combined_5 <-ny_census_combined_4 |> 
    left_join(property_tract_count, by="GEOID")


```


```{r}

saveRDS(ny_census_combined_5, "ny_census_combined_5.rds")


```



```{r}



ny_census_combined_5 <- readRDS("ny_census_combined_5.rds")


```


```{r}


ny_census_combined_5 |> 
glimpse()


```


```{r}


ny_census_combined_5 |> 
    summary()



```


```{r}



class(ny_census_combined_5)



```


```{r}


st_crs(ny_census_combined_5)

```


From the first line of our output we can see the coordinate reference system(crs) is EPSG:5030.  This is a projected crs.  A projected crs creates a map with a shpe that is a 2D representation mathematically adapted from a 3D surface.  It'a measured in linear units like meters, feet, or miles.  

The leaflet library requrirers an unprojected crs.  An unprojected crs represents the Earth as a sphere. It's measured in angular units like decimal degrees.  We will transform the ny_census_combined_5 dataset to an unprojected crs using the function st_transform from the SF library. 


```{r}

ny_census4326 <- ny_census_combined_5 |> 
    st_transform(crs=4326)


```

```{r}



saveRDS(ny_census4326 , "ny_census4326.rds")


```

```{r}

ny_census4326  <- readRDS("ny_census4326.rds")


```



```{r}




st_crs(ny_census4326)


```

## Population Density


```{r}


ny_census4326 |> 
    glimpse()

```


```{r}



ny_census4326 |> 
    filter(county_name == "Queens County" & census_tract == "Census Tract 663.01")


```



The crs is now EPSG:4326, which is unprojected. 




```{r}



bins<-c(0,8000,16000,32000,48000,60000,90000,Inf)


```

```{r}



pal<-colorBin(palette="YlOrRd",
            domain=ny_census_combined_5$pop_density,bins=bins)

```

```{r}



mytext <- paste(
  "Census Tract: ", ny_census4326$census_tract,"<br/>",
  "Population: ", ny_census4326$total_population, "<br/>",
  "County: ", ny_census4326$county_name,"<br/>", 
  "Properties For Sale: ", ny_census4326$tract_property_count, "<br/>",
    sep="") %>%
  lapply(htmltools::HTML)



```


```{r}



map_tract_pop_density_plt<- ny_census4326 %>% leaflet() %>% 
  addTiles() %>% 
  setView(lng = -73.87478, lat = 40.72395, zoom = 11.4) |> 
  #addProviderTiles(providers$CartoDB) %>% 
  addPolygons(
 fillColor=~pal(pop_density),
 weight=2,
 opacity=1,
 color= "white",
 label=~mytext,
 dashArray = "3",
 fillOpacity=0.7,
 highlightOptions = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.7,
    bringToFront = TRUE))


```





```{r}



tract_map_pop_density<-map_tract_pop_density_plt %>% 
  addLegend(pal = pal, values = ~pop_density, opacity = 0.7, title = "Census Tract Population Density",
  position = "bottomright")


```


```{r}


tract_map_pop_density


```


Zooming in on our map we see most census tracts with population density 90,000 or above mostly located in New York County (manhatten).  We do observe some census tracts in Queens County and Bronx County that have population densities 90,000 or above. 

## Median Home Value Census Tract


We'll creare a median property price by tract.  



```{r}


median_tract_price <- ny_housing_census_2 |>
    group_by(GEOID) |> 
    summarize(median_tract_property_price = median(PRICE)) |> 
    ungroup() 
   



```


Convert median_tract_propertyt_price to integer. 



```{r}


median_tract_price <- median_tract_price |> 
    mutate(median_tract_property_price = as.integer(median_tract_property_price))


```



```{r}



saveRDS(median_tract_price, "median_tract_price.rds")

```


```{r}


median_tract_price <- readRDS("median_tract_price.rds")


```


```{r}

median_tract_price |> head()


```


```{r}

median_tract_price |> glimpse()


```



Our media_tract_price dataframe will be joined to our unprojected combined data. 





```{r}


ny_combined_census4326 <- ny_census4326 |> 
    left_join(median_tract_price, join_by(GEOID))


```



```{r}


saveRDS(ny_combined_census4326, "ny_combined_census4326.rds")


```




```{r}


ny_combined_census4326 <- readRDS("ny_combined_census4326.rds")


```


```{r}


st_crs(ny_combined_census4326)


```


```{r}

ny_combined_census4326 |> head()


```


```{r}


ny_combined_census4326 |> glimpse()

```

```{r}

ny_combined_census4326 |> 
    summary()


```

Now that we have addeed our median property price feature we can build our map. 


```{r}



bins<-c(0,115000,300000,600000,900000,1200000,1500000, 1800000,Inf)



```


```{r}


pal<-colorBin(palette="YlOrRd",
            domain=ny_combined_census4326$median_value,bins=bins)


```


```{r}




mytext <- paste(
  "Census Tract: ", ny_combined_census4326$census_tract,"<br/>",
  "County: ", ny_combined_census4326$county_name,"<br/>",
  "Median Property Price: ", ny_combined_census4326$median_tract_property_price, "<br/>",
  "Median Structure Age: ", ny_combined_census4326$median_structure_age,"<br/>", 
  "Percent Owner Occupied: ", ny_combined_census4326$percent_ooh, "<br/>",
  "Median Age: ", ny_combined_census4326$median_age, "<br/>",
  "Median Income: ", ny_combined_census4326$median_income, "<br/>",
    sep="") %>%
  lapply(htmltools::HTML)



```



```{r}


map_tract_median_value_plt<- ny_combined_census4326 %>% leaflet() %>% 
  addTiles() %>% 
  setView(lng = -73.87478, lat = 40.72395, zoom = 11.4) |> 
  #addProviderTiles(providers$CartoDB) %>% 
  addPolygons(
 fillColor=~pal(median_value),
 weight=2,
 opacity=1,
 color= "white",
 label=~mytext,
 dashArray = "3",
 fillOpacity=0.7,
 highlightOptions = highlightOptions(
    weight = 5,
    color = "#666",
    dashArray = "",
    fillOpacity = 0.7,
    bringToFront = TRUE))



```


```{r}




tract_map_med_value<-map_tract_median_value_plt %>% 
  addLegend(pal = pal, values = ~median_value, opacity = 0.7, title = "Census Tract Median Home Value",
  position = "bottomright")


```


```{r}


tract_map_med_value


```


# Model


This is rstanarm version 2.32.2
- See https://mc-stan.org/rstanarm/articles/priors for changes to default priors!
- Default priors may change, so it's safest to specify priors, even if equivalent to the defaults.
- For execution on a local, multicore CPU with excess RAM we recommend calling
  options(mc.cores = parallel::detectCores())



```{r}


load_variables(2020,"acs5")



```




# Models

## Linear Regression


Our first linear regression model will include only predictor features from our original New York Housing dataset. 



```{r}




formula_1 <- "log(PRICE) ~ TYPE + BEDS + BATH + log(PROPERTYSQFT )"



```


```{r}

options(scipen = 999)


```



```{r}


model_1 <- lm(formula = formula_1, data = ny_census_combined_4)


```



```{r}


summary(model_1)


```


Our model has an R-squared of 0.518 which places it as an average model.  All of the predictor features are significant.  Let's try and improve our model by adding the census data features as predictors. 





```{r}



formula_2 <- "log(PRICE) ~ TYPE + BEDS + BATH + log(PROPERTYSQFT) + median_value + total_population + median_age + median_income + pct_college + pct_foreign_born + pct_white + percent_ooh + pop_density + median_structure_age"


```


```{r}


model_2 <- lm(formula = formula_2, data = ny_census_combined_4)


```



```{r}


summary(model_2)

```

The R squared for model 2 improved to 0.72.  All but four features are significant.  The BEDS feature, which was significant in model 1 is now insignificant.  







```{r}


library(corrr)


```


```{r}



ny_housing_estimates <- ny_census_combined_4 |> 
    select(-GEOID, -census_tract, -county_name, -median_year_built, -TYPE, -PRICE, -LATITUDE, -LONGITUDE, -county, -property_id) |> 
    st_drop_geometry()



```



```{r}


correlations <- correlate(ny_housing_estimates, method = "pearson")


```


```{r}


network_plot(correlations)


```




## Mixed Model



```{r}

scale()


```




```{r}




formula_3 <- "log(PRICE) ~ TYPE + BEDS + BATH + log(PROPERTYSQFT + (1 | GEOID))"


```


```{r}


model_3 <- lmer(log(PRICE) ~ TYPE + BEDS + BATH + log(PROPERTYSQFT) + (1 | GEOID), data = ny_census_combined_4)


```



```{r}


model_3 <- lmer(PRICE ~  (1 | GEOID), data = ny_census_combined_4)



```


```{r}


library(lmerTest)


```



```{r}


model_4 <- as_lmerModLmerTest()

```