---
title: "Analysis of the New York Housing Market"
format: html
---


```{r}

library(dplyr)
library(ggplot2)
library(plotly)
library(tidyr)
library(hexbin)
library(RColorBrewer)
library(forcats)
library(crosstalk)
library(ggiraph)

library(survey)

library(srvyr)

library(tidycensus)

library(sf)

library(jsonlite)

library(tigris)

library(stringr)

library(leaflet)
library(htmlwidgets)
library(htmltools)
library(gt)
library(DT)
library(GGally)


```




# New York Housing data Review


```{r}


saveRDS(ny_housing, "ny_housing.rds")


```


```{r}


ny_housing<-readRDS("ny_housing.rds")

```



```{r}



ny_housing |> glimpse()


```



```{r}



ny_housing |> summary()


```



```{r}

ny_housing |> 
    select(STATE, MAIN_ADDRESS, FORMATTED_ADDRESS, LOCALITY) |> 
    head(20)


```




The address fields are not consistant on the city or town names.  In some instances the city name is actually a neighborhood within a city or New York City borough. 

We will use the stringr package and regular expressions to extract the zip code which then can be used to determine the city. 





```{r}


ny_housing_2<-ny_housing |> 
    mutate(zip_cde=str_extract(MAIN_ADDRESS,"\\d{5}"))




```



```{r}


saveRDS(ny_housing_2,"ny_housing_2.rds")


```



```{r}


ny_housing_2 <- readRDS("ny_housing_2.rds")




```


```{r}


ny_housing_2 |> 
    select(zip_cde) |> 
    head(20)



```




```{r}


glimpse(ny_housing_2)


```






```{r}

ny_housing_2 |> 
    head(20)


```





```{r}






```

```{r}







```









```{r}






```












```{r}



Sys.getenv("CENSUS_API_KEY")

```



```{r}



census_2020<-load_variables(2020,"pl",cache=TRUE)


```

## Longitude/Latitude 


For our analysis we will require census tract codes that match our NY housing data. We will use the longitude and latitude columns in the dataset for this.  The first step is to create a data frame which includes only the latitude and longitude columns. 

```{r}



long_lat <- ny_housing_2 |> 
    select(LATITUDE, LONGITUDE)


```


```{r}


long_lat |> glimpse()



```



Now that we have our long_lat data frame we will use the call_geolocator_latlon function from the Tigris library.  The function submits a request to the census bureau api with a new column of census_code specified. 

```{r}



long_lat$census_code <- apply(long_lat, 1, function(row) call_geolocator_latlon(row['LATITUDE'], row['LONGITUDE']))




```


```{r}


long_lat_data <-long_lat



```


```{r}

saveRDS(long_lat_data,"long_lat_data.rds")



```



```{r}


long_lat_data <-readRDS("long_lat_data.rds")

```


We can observe that the data frame returned from the cenus api now includes the cenus_code column. 


```{r}



long_lat_data |> head()


```

```{r}


long_lat_data |> glimpse()


```


The 15 digit code returnd  from the census api is made up of serveral codes put together.  The first two digits are the state code.  The next three digits are the county code. The next six digits are the census tract code.  The last four digits are the census block code. 


```{r}



ny_tracts <- tracts("NY", year = 2024)


```

```{r}


class(ny_tracts)



```

```{r}


ny_tracts |>  head()


```


```{r}


census_codes<-long_lat_data



```



We will use the substr function to break up the census code into its individual components.  THe census tract will first be extracted. 


```{r}



census_codes$census_tract<-substr(census_codes$census_code,6,11)


```



Three digit county code

```{r}



census_codes$county <-substr(census_codes$census_code,3,5)



```



four digit census block code
 


```{r}

census_codes$block <-substr(census_codes$census_code,12,15)


```

GEOID column



```{r}



census_codes$GEOID <-substr(census_codes$census_code,1,11)


```



```{r}


saveRDS(census_codes, "census_codes.rds")



```


```{r}


census_codes <- readRDS("census_codes.rds")


```


```{r}



census_codes |> head()


```

```{r}


census_codes |>  glimpse()



```



The output displays our added census tract, count, block codes, and GEOID. 






```{r}


census_geo <- census_codes |> 
    select(-LATITUDE,-LONGITUDE)


```


```{r}



saveRDS(census_geo,"census_geo.rds")


```


```{r}


census_geo <- readRDS("census_geo.rds")

```

```{r}

census_geo |> head()



```

```{r}


census_geo |>  glimpse()

```




```{r}


ny_housing_census <- cbind(ny_housing_2, census_geo)


```



```{r}



ny_housing_census <- ny_housing_census |> 
    select(GEOID,census_code,TYPE, PRICE, BEDS, BATH, PROPERTYSQFT, LATITUDE, LONGITUDE, zip_cde, census_tract, county, block)


```



```{r}



saveRDS(ny_housing_census, "ny_housing_census.rds")


```


```{r}


ny_housing_census <- readRDS("ny_housing_census.rds")


```

```{r}

ny_housing_census |> head()


```




```{r}


ny_housing_census |> glimpse()



```


```{r}



ny_housing_census |> summary()

```



```{r}



ny_housing_census |> 
    count(TYPE) |> 
    arrange(desc(n))



```


For TYPE, certain categories (Pending, Contingent, For Sale, Foreclsure, Coming Soon) offer little information on what the property is. Additionally, "Mobile  house for sale" has a count of only one. These categories will be removed.  


```{r}


ny_housing_census_2 <- ny_housing_census |> 
    filter(TYPE %in% c("Co-op for sale", "House for sale", "Condo for sale", "Multi-family home for sale", "Townhouse for sale", "Land for sale", "Condop for sale"))


```



We will create artifical id's for each property.  This will be useful during our modeling. 


```{r}



ny_housing_census_2 <- ny_housing_census_2 |> 
    mutate(property_id = row_number())


```


The TYPE category "Condop for sale" will be merged into "Condo for sale" 


```{r}



ny_housing_census_2 <- ny_housing_census_2 |> 
    mutate(TYPE = recode(TYPE, "Condop for sale" = "Condo for sale"))




```


```{r}


ny_housing_census_2 <- ny_housing_census_2 |> 
    mutate(BATH = round(BATH,0)) |> 
    mutate(PROPERTYSQFT = round(PROPERTYSQFT,0))


```



```{r}


ny_housing_census_2 <- ny_housing_census_2 |> 
    select(-census_tract)



```



```{r}


ny_housing_census_2 <- ny_housing_census_2 |> 
    mutate(property_id = as.character(property_id))

```


```{r}



saveRDS(ny_housing_census_2, "ny_housing_census_2.rds")


```


```{r}


ny_housing_census_2 <- readRDS("ny_housing_census_2.rds")


```


```{r}


ny_housing_census_2 |> 
    count(TYPE) |> 
    arrange(desc(n))

```

The feature TYPE has been reduced to six categories. 



```{r}


ny_housing_census_2 |> 
    count(TYPE) |> 
    mutate(perc = round(n/sum(n),2)) |> 
    arrange(desc(perc))



```

Above is the ercentage breakdown of propery type.  type. 





```{r}



ny_housing_census_2 |> 
    select(property_id) |>
     head()


```






```{r}


dim(ny_housing_census_2)



```

Our dataset now has 4,433 observations. 


```{r}



```





```{r}


ny_housing_census_2 |> 
    count(county) |> 
    arrange(desc(n))


```


There are five counties in our data that have properties for sale. 




```{r}



ny_housing_census_2 |> 
    count(TYPE, GEOID) |>
    slice_max(n,n=10) |> 
    arrange(desc(n))



```

The top 10 tracts by count of property for sale. 


```{r}


options(scipen = 999)


```


```{r}

ny_housing_census_2 |> 
    select(PRICE, BEDS, BATH, PROPERTYSQFT) |> 
    summary()


```

## Visualization

```{r}


sqr_ft_box<-ny_housing_census_2 %>% 
  ggplot(aes(x=TYPE, y=PROPERTYSQFT, fill=TYPE))+
  geom_boxplot()+
  ggtitle("Property Square Feet by Property Type")+
  theme(legend.position = "none",plot.title = element_text(color="blue", size=14, face="bold.italic", hjust=0.5))+
  theme(axis.ticks.x=element_blank(),legend.position = "none", axis.text.x=element_text(angle=70, hjust=1),
        axis.title.x=element_blank(),
    axis.title.y=element_blank())


```

```{r}

sqr_ft_box

```

```{r}


bed_box<-ny_housing_census_2 %>% 
  ggplot(aes(x=TYPE, y=BEDS, fill=TYPE))+
  geom_boxplot()+
  ggtitle("Number of Bed Rooms  by Property Type")+
  theme(legend.position = "none",plot.title = element_text(color="blue", size=14, face="bold.italic", hjust=0.5))+
  theme(axis.ticks.x=element_blank(),legend.position = "none", axis.text.x=element_text(angle=70, hjust=1),
        axis.title.x=element_blank(),
    axis.title.y=element_blank())


```




```{r}

bed_box

```



```{r}



bath_box<-ny_housing_census_2 %>% 
  ggplot(aes(x=TYPE, y=BATH, fill=TYPE))+
  geom_boxplot()+
  ggtitle("Number of Bath Rooms  by Property Type")+
  theme(legend.position = "none",plot.title = element_text(color="blue", size=14, face="bold.italic", hjust=0.5))+
  theme(axis.ticks.x=element_blank(),legend.position = "none", axis.text.x=element_text(angle=70, hjust=1),
        axis.title.x=element_blank(),
    axis.title.y=element_blank())


```






```{r}

bath_box


```





```{r}

corr_plot <- ny_housing_census_2 |> 
    select(PRICE, PROPERTYSQFT, BEDS, BATH) |> 
    ggscatmat(corMethod = "pearson", alpha = 0.2)





```


```{r}




corr_plot




```








```{r}


write.csv(ny_housing_census, "ny_housing_census.csv")



```





# Census 




```{r}



load_variables(2024, "acs5")


```


## Tract Demographic Data


Requesting census data from different tables can cause errors. To prevent this, we will request data separatley for the two tables. 


```{r}


variables_to_get_1 <- c(
    median_value = "B25077_001",
    total_population = "B01003_001",
    median_age = "B01002_001",
    median_year_built = "B25037_001"
   

) 






```


```{r}




variables_to_get_2 <- c(
    median_income = "DP03_0062",
    pct_college = "DP02_0068P",
    pct_foreign_born = "DP02_0094P",
    pct_white = "DP05_0077P",
    percent_ooh = "DP04_0046P"

) 


```


```{r}


ny_tract_info_1 <- get_acs(
    geography = "tract",
    variables = variables_to_get_1,
    state = "NY",
    #geometry = TRUE,
    output="wide",
    year = 2024
) #|> 
    #select(-NAME)



```


```{r}



ny_tract_info_1 |> head()



```

```{r}



ny_tract_info_1 |> glimpse()


```





```{r}


saveRDS(ny_tract_info_1,"ny_tract_info_1.rds")



```

```{r}



ny_tract_info_1 <- readRDS("ny_tract_info_1.rds")


```

```{r}


ny_tract_info_2 <- get_acs(
    geography = "tract",
    variables = variables_to_get_2,
    state = "NY",
    #geometry = TRUE,
    output="wide",
    year = 2024
) #|> 
    #select(-NAME)




```



```{r}


ny_tract_info_2 <- ny_tract_info_2 |> 
    select(-geometry)



```


```{r}



saveRDS(ny_tract_info_2,"ny_tract_info_2.rds")



```






```{r}



ny_tract_info_2 <- readRDS("ny_tract_info_2.rds")


```


```{r}

ny_tract_info_2 |> glimpse()



```


```{r}

ny_tract_geometry <- ny_tract_info_1 |> 
    select(GEOID, geometry)


```


```{r}


ny_tract_geometry |> glimpse()


```


```{r}


saveRDS(ny_tract_geometry, "ny_tract_geometry.rds")


```


```{r}



ny_tract_geometry <- readRDS("ny_tract_geometry.rds")


```


```{r}


ny_tract_geometry |> head()


```


```{r}



class(ny_tract_geometry)


```

```{r}

ny_tract_cls <- ny_tract_geometry |> 
    st_transform(5070) 


```


```{r}


saveRDS(ny_tract_cls, "ny_tract_cls.rds")



```


```{r}


ny_tract_cls <- readRDS("ny_tract_cls.rds")


```


```{r}


class(ny_tract_cls)


```

```{r}



ny_tract_cls |> head()



```


```{r}


ny_tract_data <- ny_tract_info_1 |> 
    full_join(ny_tract_info_2, by="GEOID")


```




```{r}







```


```{r}


ny_tract_data |> glimpse()


```


```{r}


ny_tract_data <-ny_tract_data |> 
    select(-NAME.y)


```

```{r}

ny_tract_data <- ny_tract_data |> 
    rename(NAME=NAME.x)

```



```{r}


saveRDS(ny_tract_data, "ny_tract_data.rds")




```



```{r}



ny_tract_data <-readRDS("ny_tract_data.rds")



```



```{r}



ny_tract_data |> glimpse()

```


```{r}



ny_tract_data_2 <- ny_tract_data |> 
    separate(NAME, into=c("census_tract", "county", "state"), sep="; ")


```


```{r}



ny_tract_data_2 |> glimpse()



```




```{r}



ny_tract_data_2 <- ny_tract_data_2 |> 
    select(-state)



```

```{r}





```



```{r}

saveRDS(ny_tract_data_2, "ny_tract_data_2.rds")


```


```{r}



ny_tract_data_2 <-readRDS("ny_tract_data_2.rds")


```

```{r}

ny_tract_data_2 |> summary()



```


```{r}



ny_tract_data_2 |> glimpse()


```



```{r}



ny_tract_geom<-ny_tract_cls |> 
    full_join(ny_tract_data_2, by = "GEOID")


```


```{r}

saveRDS(ny_tract_geom,"ny_tract_geom.rds")


```


```{r}


ny_tract_geom <- readRDS("ny_tract_geom.rds")


```

```{r}

ny_tract_geom |> glimpse()


```

```{r}


ny_tract_geom |> summary()


```



```{r}


#ny_tract_geom <- st_join(ny_tract_data,ny_tract_geometry)


```

## Feature Engineering

```{r}


library(units)


```


```{r}


ny_tract_geom_2 <- ny_tract_geom |> 
    mutate(pop_density = as.numeric(set_units(total_populationE / st_area(geometry),"1/km2")),
           median_structure_age = 2022 - median_year_builtE) |> 
    select(!ends_with("M")) |> 
    rename_with(.fn = ~str_remove(.x, "E$"))



```



```{r}



ny_tract_geom_2 <- ny_tract_geom_2 |> 
    mutate(pop_density = round(pop_density,2))



```


```{r}


ny_tract_geom_2 <- ny_tract_geom_2 |> 
    rename(county_name = county)



```



```{r}


saveRDS(ny_tract_geom_2 , "ny_tract_geom_2.rds")


```



```{r}



ny_tract_geom_2  <- readRDS("ny_tract_geom_2.rds")


```



```{r}


ny_tract_geom_2 |> head(20)


```


```{r}



ny_tract_geom_2 |> 
    summary()


```



```{r}


ny_tract_geom_2 |> 
    filter(census_tract == "Census Tract 102") |> 
    head()


```




```{r}

ny_tract_geom_2 |> 
    select(census_tract) |> 
    count(census_tract) |> 
    arrange(desc(n))


```






```{r}



ny_tract_geom_2 |> 
    glimpse()




```

## Join Housing and Tract Data



```{r}





ny_census_combined <- ny_tract_geom_2 |> 
    inner_join(ny_housing_census_2, join_by(GEOID)) |> 
    select(-census_code, -LATITUDE, -LONGITUDE, -zip_cde, -block)



```


```{r}


saveRDS(ny_census_combined, "ny_census_combined.rds")

```


```{r}

ny_census_combined <- readRDS("ny_census_combined.rds")


```




# Data Cleaning

## Missing Values


```{r}



ny_census_combined |> glimpse()

```

```{r}

ny_census_combined |>  summary()


```



There are a number of features with missing values.  Lets inspect these missing values. 



### Median year built

```{r}



ny_census_combined |> 
    filter(is.na(median_year_built)) |> 
    summary()



```


```{r}




ny_census_combined |> 
    filter(is.na(median_year_built)) |> 
    head()



```

There are three Census tracts which are missing demographic data.  We will remove these missing values from the dataset. 



```{r}



ny_census_combined_2 <- ny_census_combined |> 
    filter(!is.na(median_year_built))



```


```{r}

saveRDS(ny_census_combined_2, "ny_census_combined_2.rds")


```

```{r}


ny_census_combined_2 <- readRDS("ny_census_combined_2.rds")


```




```{r}


ny_census_combined_2 |> 
    filter(is.na(median_year_built)) |> 
    summary()


```

The three census tracts with missing demographic data have been removed. 


### Medium Income


```{r}




ny_census_combined_2 |> 
    summary()




```


There are missing values for median_income.  

```{r}


ny_census_combined_2 |> 
    filter(is.na(median_income)) |> 
    head(17)


```


```{r}

ny_census_combined_2 |> 
    filter(is.na(median_income)) |> 
    tail(10)


```

The tracts with missing median_income values have pct_white populations that are low or zero.  AS such it would be best for our analyis to retain these rows.  We will replace missing values with the median of the tract median_income.   


```{r}






```

Census tracts 102 and 96 within New York County

We'll use median income of all tracts within the county the tract witgh the missing values reside in. 

```{r}


ny_tract_geom_2 |> 
    filter(county_name == "New York County") |> 
    filter(!is.na(median_income)) |> 
    summarise(median_tract_income = median(median_income))



```


```{r}



ny_census_combined_3 <- ny_census_combined_2 |> 
    mutate(median_income = ifelse(county_name == "New York County" & is.na(median_income), 121760.5,median_income ))


```



```{r}


ny_census_combined_3 |> 
    filter(is.na(median_income)) |> 
    head(17)


```

As New York County includes tracts 102 and 96, we can observe those those tracts no longer have missing values for median income. 






Census Tract 226.02 Richmond County

```{r}

ny_tract_geom_2 |> 
    filter(county_name == "Richmond County") |> 
    filter(!is.na(median_income)) |> 
    summarise(median_tract_income = median(median_income))



```





census tract 226.02



```{r}

ny_census_combined_3 |> 
    filter(census_tract == "Census Tract 102") |> 
    filter(is.na(median_income)) 


```


```{r}


ny_census_combined_3 <- ny_census_combined_3 |> 
    mutate(median_income = ifelse(county_name == "Richmond County" & is.na(median_income), 99305,median_income ))

```



Census Tract 932  Kings County 

```{r}



ny_tract_geom_2 |> 
    filter(county_name == "Kings County") |> 
    filter(!is.na(median_income)) |> 
    summarise(median_tract_income = median(median_income))



```



```{r}



ny_census_combined_3 <- ny_census_combined_3 |> 
    mutate(median_income = ifelse(county_name == "Kings County" & is.na(median_income), 79773,median_income ))


```


Census Tract 231  Bronx County



```{r}


ny_tract_geom_2 |> 
    filter(county_name == "Bronx County") |> 
    filter(!is.na(median_income)) |> 
    summarise(median_tract_income = median(median_income))


```


```{r}


ny_census_combined_3 <- ny_census_combined_3 |> 
    mutate(median_income = ifelse(county_name == "Bronx County" & is.na(median_income), 50243,median_income ))


```


Census Tract 536.01 Queens County

```{r}


ny_tract_geom_2 |> 
    filter(county_name == "Queens County") |> 
    filter(!is.na(median_income)) |> 
    summarise(median_tract_income = median(median_income))

```


```{r}


ny_census_combined_3 <- ny_census_combined_3 |> 
    mutate(median_income = ifelse(county_name == "Queens County" & is.na(median_income), 88882,median_income ))


```


```{r}


saveRDS(ny_census_combined_3, "ny_census_combined_3.rds")




```



```{r}

ny_census_combined_3 <- readRDS("ny_census_combined_3.rds")


```


```{r}



ny_census_combined_3 |> 
    filter(is.na(median_income)) |> 
    head()



```

```{r}


ny_census_combined_3 |> 
    summary()


```






All missing values for median income have been filled in (impouted).  The last feature to inspect is median_value. 

### Median Value


```{r}




ny_census_combined_3 |> 
    filter(is.na(median_value)) |> 
    head(10)


```


```{r}



ny_census_combined_3 |> 
    filter(is.na(median_value)) |> 
    tail(10)


```





 Census Tracts 391, 215.02 , 266.01, 375.04, 173, 419 Bronx County








```{r}


ny_tract_geom_2 |> 
    filter(county_name == "Bronx County") |> 
    filter(!is.na(median_value)) |> 
    summarise(median_tract_value = median(median_value))



```




```{r}



ny_census_combined_4 <- ny_census_combined_3 |> 
    mutate(median_value = ifelse(county_name == "Bronx County" & is.na(median_value), 607050,median_value ))



```




Census Tracts 505, 596, 575 Queens County





```{r}


ny_tract_geom_2 |> 
    filter(county_name == "Kings County") |> 
    filter(!is.na(median_value)) |> 
    summarise(median_tract_value = median(median_value))


```



```{r}


ny_census_combined_4 <- ny_census_combined_4 |> 
    mutate(median_value = ifelse(county_name == "Kings County" & is.na(median_value), 956900,median_value ))


```



Census Tracts 96, 172, 99.0293 New York County


```{r}


ny_tract_geom_2 |> 
    filter(county_name == "New York County") |> 
    filter(!is.na(median_value)) |> 
    summarise(median_tract_value = median(median_value))



```

```{r}


ny_census_combined_4 <- ny_census_combined_4 |> 
    mutate(median_value = ifelse(county_name == "New York County" & is.na(median_value), 1078250,median_value ))

```

Tracts 85, 938, 83, 1227.02, 1133, 779.08, 155, 240 Queens County


```{r}


ny_tract_geom_2 |> 
    filter(county_name == "Queens County") |> 
    filter(!is.na(median_value)) |> 
    summarise(median_tract_value = median(median_value))

```



```{r}


ny_census_combined_4 <- ny_census_combined_4 |> 
    mutate(median_value = ifelse(county_name == "Queens County" & is.na(median_value), 746350,median_value ))



```



```{r}


ny_census_combined_4 <- ny_census_combined_4 |> 
    mutate(property_id = as.character(property_id))

```


```{r}


saveRDS(ny_census_combined_4 , "ny_census_combined_4.rds")


```


```{r}


ny_census_combined_4  <- readRDS("ny_census_combined_4.rds")


```

All missing values have been removed or imputed. 


```{r}


ny_census_combined_4 |> 
    summary()


```


```{r}

ny_census_combined_4 |> glimpse()


```

# Visualization



```{r}

ny_tracts <- tracts(state="NY", cb=FALSE, year=2024)


```



```{r}

saveRDS(ny_tracts, "ny_tracts.rds")


```



```{r}


ny_tracts <- readRDS("ny_tracts.rds")


```



```{r}



ny_census_combined_4 |> 
    group_by(property_id,county_name, census_tract) |> 
    count(property_id) |>
    slice_max(n,n=10) |> 
    ungroup()
    #print(n=50)


```


```{r}



ny_tracts <- readRDS("ny_tracts.rds")


```


```{r}


ny_tracts_plot <- ggplot(ny_tracts)+
    geom_sf()+
    theme_void()



```


```{r}






```


```{r}



ny_tracts <- tracts(state="NY", cb=FALSE, year=2024)


```


We're going to create a map displaying the median home value of the census tract. First, we are going to create a feature for the property count in each census tract. This will be used for hovering data on our map.   

```{r}


property_tract_count<-ny_housing_census_2 |> 
    count(GEOID, name = "tract_property_count")
    


```



```{r}

saveRDS(property_tract_count, "property_tract_count.rds")


```


```{r}

property_tract_count <- readRDS("property_tract_count.rds")


```



```{r}


property_tract_count |> 
    glimpse()


```

```{r}


property_tract_count |> 
    slice_max(tract_property_count, n=20) 
    
   


```


```{r}

property_tract_count |> 
    summary()



```


```{r}


class(property_tract_count)



```


We will now add our tract property count feature to our combined data. 


```{r}


ny_census_combined_5 <-ny_census_combined_4 |> 
    left_join(property_tract_count, by="GEOID")


```


```{r}

saveRDS(ny_census_combined_5, "ny_census_combined_5.rds")


```



```{r}



ny_census_combined_5 <- readRDS("ny_census_combined_5.rds")


```


```{r}


ny_census_combined_5 |> 
glimpse()


```


```{r}


ny_census_combined_5 |> 
    summary()



```


```{r}



class(ny_census_combined_5)



```


```{r}


load_variables(2020,"acs5")



```